Why there are more species in several small patches vs few large patches:

### 2024-07-01 RunSims.Rmd
New simulations no-selection with habitat destruction of 0.4 and 0.8

>>>>

## Lambda = 1-4 NoSelection = Neutral  Regular prog frag 0.4-0.8

```{r SLOSS_lambda14_pf48_regular,echo=FALSE,message=FALSE,warning=FALSE,fig.align='center',eval=TRUE}

# pf = probability of fragmentation, hps = habitat patch size , dd=dispersal distance 
# hps = habitat patch size
# dd = dispersal distance 
# brb = bird rate birds 
nl@experiment <- experiment(expname="NoSelection_pf04-08_hps3-61_dd1-3_brb1-4_regular",
                            outpath=outpath,
                            repetition=1,
                            tickmetrics="false",
                            idsetup="setup",
                            idgo="go",
                            runtime=0,
                            metrics=c("birds-at-deforestation", "calc-number-of-species","calc-shannon-diversity","count birds", "mean-free-path","find-cluster-sizes"),
                            variables = list("birds-dispersal-distance" = list(values=c(1.1, 1.2, 1.5, 2, 3)),
                                              "habitat-patch-size" = list(values=c(3, 9, 29, 61)),
                                               "birth-rate-birds" = list(values=c(1.7, 2, 3, 4)),
                                             "prob-frag" =  list(values=c(0.4,0.8))
                                             ),
                            constants = list("world-width" = 199,
                                             "world-height" = 199,
                                             "initial-population" = 0,
                                             "max-birds-species" = 100,
                                             "death-rate-birds" = 1,
                                             "migration-rate-birds" = 0.0001,
                                             "Video" = "false",
                                             "Deforestation-at-time" = 200,
                                             "end-time" = 1500,
                                             "Migration0-at-deforestation" = "false",
                                             "deforestation-type" = "\"regular\"",
                                             "birds-behavior" = "\"NoSelection\"",
                                             "replacement-rate" = 0.0
                                             ))

### 2024-07-01 ms_SLOSS.Rmd
Add Format files
New simulations no-selection with habitat destruction of 0.4 and 0.8

---
title: "Why there are more species in several small patches vs few large patches:
  a multi-species modeling approach"
bibliography: sloss_bib.bib
csl: "ecology-letters.csl"

output:
  pdf_document: 
    includes:
      in_header: margins.sty
---

@@

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. The policies of many conservation agencies are still prioritizing the preservation of a single large habitat area (SL) vs. several small areas of the same size (SS), despite empirical support favouring SS. However, to date, most theories regarding this issue were based on single-species models. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine its causes. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is insufficient to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. The model predicts that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Habitat fragmentation, biodiversity, spatially explicit models, multi-species models, neutral theory, competitive theory, habitat selection, dispersal, extinction, reproduction.

>>>>

---
title: "Why there are more species in several small patches vs few large patches:
  a multi-species modeling approach"
bibliography: sloss_bib.bib
csl: "functional-ecology.csl"
output:
  pdf_document:
    includes:
      in_header: margins.sty
  word_document: default
---

@@

1. As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. The policies of many conservation agencies are still prioritizing the preservation of a single large habitat area (SL) vs. several small areas of the same size (SS), despite empirical support favouring SS. However, to date, most theories regarding this issue were based on single-species models.

2. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine its causes. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes.

3. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is insufficient to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species.

4. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. The model predicts that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Habitat fragmentation, biodiversity, spatially explicit models, multi-species models, neutral theory.
